### 2015-06-06 chapters/ipa_meets_unicode.tex
restoring pitfall on digraphs

Some other homoglyphs related to encoding IPA in the Unicode Standard are:

\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list. An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at \uni{0027}, although the \textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{} \textsc{right single quotation mark} at \uni{2019}. Yet the glottal stop/glottalization/ejective marker is another completely different character, the \textsc{modifier letter apostrophe} at \uni{02BC}, but unfortunately looks mostly highly similar to \uni{2019}. 
	\item There is ambiguous encoding of IPA segments within the Unicode Standard. An example is the \uni{02C1} \textsc{modifier letter reversed glottal stop} vs the \uni{02E4} \textsc{modifier letter small reversed glottal stop} . Both are denoted in the Unicode Standard as the \textsc{pharyngealized diacritic} and both appear in various resources representing phonetic data online. This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at \uni{01C3} is labeled \textsc{latin letter retroflex click}, but in IPA is an alveolar or postalveolar click (not retroflex). This naming is probably best seen as a simple error in the Unicode Standard. Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g. <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7}). Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding. However, their usage is discouraged by the Unicode core specification. Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as \textsc{ligatures} or \textsc{digraphs} (confusingly both names appear interchangeably). Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself. In the context of IPA, characters like the following ligatures should thus \emph{not} be used. Instead a combination of two characters is preferred: 
	\item <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} (use <d> + <z> instead) 
	\item <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7} (use <t> + <ʃ> instead) 
	\item <ʩ> \textsc{latin small letter feng digraph} at \uni{02A9} (use <f> + <ŋ> instead) 
\end{itemize}

@@

Reversely, <ç> \textsc{latin small letter c with cedilla} at \uni{00E7} will be
decomposed into <c> \textsc{latin small letter c} at \uni{0063} with 
<\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
also if this is not intended.

>>>>

Some other homoglyphs related to encoding IPA in the Unicode Standard are:
\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode
       Standard email list. An English keyboard inputs
       \textless{}`\textgreater{} APOSTROPHE at \uni{0027}, although the
       \textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{}
       \textsc{right single quotation mark} at \uni{2019}. Yet the glottal
       stop/glottalization/ejective marker is another completely different
       character, the \textsc{modifier letter apostrophe} at \uni{02BC}, but
       unfortunately looks mostly highly similar to \uni{2019}. 
	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
       An example is the \uni{02C1} \textsc{modifier letter reversed glottal
       stop} vs the \uni{02E4} \textsc{modifier letter small reversed glottal
       stop} . Both are denoted in the Unicode Standard as the
       \textsc{pharyngealized diacritic} and both appear in various resources
       representing phonetic data online. This is thus an example for which the
       Unicode Standard does not solve the linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the
       Unicode Consortium does not match the IPA's description: in the Unicode
       Standard at \uni{01C3} is labeled \textsc{latin letter retroflex click},
       but in IPA is an alveolar or postalveolar click (not retroflex). This
       naming is probably best seen as a simple error in the Unicode Standard.
\end{itemize}

\section{Pitfall: Ligature and digraphs}
\label{pitfall-ligatures-digraphs}

One important distinction to acknowledge is the difference between multigraphs
and ligatures. Multigraphs are groups of characters (in the context of IPA e.g.
< tʃ > or < ou >) while ligatures are single characters (e.g. <ʧ> \textsc{latin
small letter tesh digraph} at \uni{02A7}). Ligatures arose in the context of
printing easier-to-read texts, and are included in the Unicode Standard for
reasons of legacy encoding. However, their usage is discouraged by the Unicode
core specification. Specifically related to IPA, various phonetic combinations
of characters (typically affricates) are available as single code-points in the
Unicode Standard, but are designated as \textsc{ligatures} or \textsc{digraphs}
(confusingly both names appear interchangeably). Such glyphs might be used by
software to produce a pleasing display, but they should not be hard-coded into
the text itself. In the context of IPA, characters like the following ligatures
should thus \emph{not} be used. Instead a combination of two characters is
preferred:

\begin{itemize} 
	\item <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} (use <d> + <z> instead) 
	\item <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7} (use <t> + <ʃ> instead) 
	\item <ʩ> \textsc{latin small letter feng digraph} at \uni{02A9} (use <f> + <ŋ> instead) 
\end{itemize}

@@

Reversely, <ç> \textsc{latin small letter c with cedilla} at \uni{00E7} will be
decomposed into <c> \textsc{latin small letter c} at \uni{0063} with 
<\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
also if such a decompotion is not intended, because it is meaningless in IPA.
